%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% This text file is part of the source of 
%%%% `Introduction to High-Performance Scientific Computing'
%%%% by Victor Eijkhout, copyright 2012-2022
%%%%
%%%% This book is distributed under a Creative Commons Attribution 3.0
%%%% Unported (CC BY 3.0) license and made possible by funding from
%%%% The Saylor Foundation \url{http://www.saylor.org}.
%%%%
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\thispagestyle{empty}
\section*{Preface}

The field of high performance scientific computing
requires, in addition to a broad of scientific knowledge and 'computing
folkore', a number of practical skills. Call it the `carpentry' aspect
of the craft of scientific computing.

As a companion to the book `Introduction to High Performance
Scientific Computing', which covers background knowledge, here is then
a set of tutorials on those practical skills that are important to
becoming a successful high performance practitioner.

The tutorials
should be done while sitting at a computer. Given the practice of
scientific computing, they have a clear Unix bias.

\begin{notlulu}
\paragraph*{\bf Public draft}

This book is 
open for comments.
What is missing or incomplete or unclear? Is material
presented in the wrong sequence? Kindly mail me with any comments you
may have.
\end{notlulu}

\begin{download}
You may have found this book in any of a number of places; the
authoritative download location is 
\url{https://theartofhpc.com/}
That page also links to \n{lulu.com} where you can get a nicely printed copy.
\end{download}
\begin{lulu}
Thank you for buying a printed copy of this book. 
You can also download a pdf 
version, as well as lecture slides and complete source code from
\url{https://theartofhpc.com/}
\end{lulu}

\bigskip
\noindent
Victor Eijkhout {\tt eijkhout@tacc.utexas.edu}\\
Research Scientist\\
Texas Advanced Computing Center\\
The University of Texas at Austin

